---
title: "SSRP 2015"
author: "Lennon Lab"
date: "Monday, July 13, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
Here, we will use R and RStudio to examine patterns of change in resource use between an ancestral strain a derived strain. We will begin by focusing on three aspects:

### 1.) Average resource use 
One of the simplest, and hence, first questions to ask is whether there are changes in average resource use and the average number of resources used. For instance, does the derived use more resources or less resources? Is the derived able to use each resource to a greater or lesser degree?

### 2.) Resource breadth vs. capacity
A second and more in-depth question is how the number of resources used relates to the degree to which an organisms typically uses resources. This may relate to the potential for the organism to use generalist (i.e., okay at using many resources, "jack of all trades") and specialist strategies (i.e., really good at using just a few resources).

### 3.) Changes in the metabolic "fingerprint"
Note how the first two questions do not actually account for *which* resources get used or how the various resources are related by a common element or nutrient. However, by combining information on which resources are used, how greatly they are used, and how the resources are related, we can obtain a metabolic "fingerprint" for the organism. In this way, we can ask whether the fingerprint has changed between the ancestral strain and the derived strain.


## 1.) SETUP
### A. Clear the Workspace and Retrieve and Set the Working Directory
We will begin our analysis by clearing our workspace of any variables, that is, leftover information from any previous work. 
We then set our 'working directory', which points R to the folder that contains our project's files (i.e., SSRP_2015).

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
rm(list = ls()) 
getwd() 
setwd("~/GitHub/SSRP_2015") 
```

### Install Packages
Next, we will import statistical packages that will allow us to analyze our data with specific tools.
We will begin by importing a package called `vegan`, which is often used by biologists and ecologists.
Let's install the `vegan` package and its dependencies. 
If you are running `vegan` for the first time, you will need to install it with the `install.packages` function. 
Otherwise, we recommend you just use the `require` function to load `vegan` and its dependencies.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
#install.packages("vegan") 
#install.packages("reshape") 
require("vegan")
require("reshape")
```

## DEFINE A FUNCTION FOR RESOURCE USE
A look at the data will reveal that among species, isolates, and replicates, that nearly all resourcess appear to be used to some degree.
Even water appears to be used, despite the absence of nutrients in it.
Consequently, we need to determine which values constitute real use.
To do this, we will simply not count any resource as being used, if the corresponding value is less than two standard deviations away from the value for water.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
cutoff <- function(avg= "avg.water", sd = "sd.water", vals){

  # should also return the error
  # mean, error, sd, should be based on resource qualified as being used

  cutoff <- 2*sd
  num <- 0
  
  for (val in vals){
    if (val >= avg+cutoff){
      num = num + 1      
      } 
    } 
  return(num)
}

```

## 2) LOADING DATA
Here, we will load the data generated from our plate reader and the Ecolog software as well as a table of names corresponding to each resource and location on the plates.
But first, we need to find the average use values for water that were recorded for each species at a given time.
We will use these values, which really do not reflect resource use, to normalize (or make comparable) the resource use values across different plates.

```{r, results = 'hide', echo=FALSE, message = FALSE, warning = FALSE}
#file.path <- "~/GitHub/SSRP_2015/data/cleaned/"
file.path <- "./data/cleaned/"

file.names <- list.files(path = file.path, all.files=FALSE, 
                    full.names=FALSE, ignore.case=FALSE, include.dirs=FALSE)

# read in ancestor files and grab average water use value 
# Length must equal number of strain IDs in folder
water.vals <- vector(mode="list", length=4) 
# create a list of unique identifiers by combining the id and duration
# All strains in the data folder must be on this line
names(water.vals) <- c('0711.24', '0711.48', '0723.48', '0703.48')


for (name in file.names){
  file.name.info <- strsplit(name, "\\.") # split file name
  sp.id <- file.name.info[[1]][2] # organism ID
  AorD <- file.name.info[[1]][3] # ancestral or derived
  duration <- file.name.info[[1]][6] # 24 or 48 hrs
  
  if (AorD == 'A'){ # if the strain is ancestral...
    new.name <- paste(sp.id, duration, sep='.')
    dat <- as.matrix(read.table(paste(file.path, name, sep='')))
    avg <- mean(dat[1,1], dat[1,5], dat[1,9])
    index <- match(new.name, names(water.vals))
    water.vals[[index]] <- avg
    }
  } 
```

Next we save the world... and normalize our data
```{r, results = 'hide', echo=FALSE, message = FALSE, warning = FALSE}
#file.path <- "~/GitHub/SSRP_2015/data/cleaned/"
file.path <- "./data/cleaned/"
file.names <- list.files(path = file.path, all.files=FALSE, 
                    full.names=FALSE, ignore.case=FALSE, include.dirs=FALSE)

file.name.data <- c()
names <- c()
saved <- c()
for (name in file.names){
  file.name.info <- strsplit(name, "\\.") # split file name
  file.name.info <- file.name.info[[1]][-7]
  
  sp.id <- file.name.info[[2]] # organism ID
  rep <- paste(file.name.info[[4]], file.name.info[[5]], sep="")
  AorD <- file.name.info[[3]] # ancestral or derived
  duration <- file.name.info[[6]] # 24 or 48 hrs
  data <- as.matrix(read.table(paste(file.path, name, sep='')))
  assign(paste(sp.id, AorD, rep, duration, sep="."), data)
  names <- c(names, sp.id)
  saved <- c(paste(sp.id, AorD, rep, duration, sep="."), saved) 
  }

strains <- levels(as.factor(names))
resource.names <- as.matrix(read.table("./ecoplate.info/resource_matrix.txt"))
mol.groups <- as.matrix(read.delim("./ecoplate.info/moleculetype_matrix.txt", header=F))
resources <- levels(as.factor(resource.names))
r.names <- as.factor(resource.names)[1:32]
c.grouping <- as.factor(mol.groups)[1:32]
group.res <- data.frame(r.names, c.grouping)[-1, ]
resources <- resources[resources != "Water"]

for (i in strains){
  data <- as.data.frame(matrix(NA, 31, 15))
  colnames(data) <- c("Resource", "Group", "Anc", "D1.A", "D1.B", "D1.C", "D2.A", "D2.B", "D2.C", "D3.A", "D3.B", "D3.C", "D1fold", "D2fold", "D3fold")
  data$Resource <- resources
  Anc <- get(saved[grep(paste(i, ".A.*.48", sep=""), saved)])
  anc.avg.water <- mean(c(Anc[1,1], Anc[1,5], Anc[1,9]))
  for (j in resources){
    data$Group[which(data$Resource == j)] <- as.character(group.res$c.grouping[which(group.res$r.names == j)])
    }
  for (k in resources){
    data$Anc[which(data$Resource  == k)] <- mean(Anc[which(resource.names == k)])
  } 
  for (l in 1:length(grep(paste(i, ".D.*.48", sep=""), saved))){
    Der <- get(saved[grep(paste(i, ".D.*.48", sep=""), saved)][l])
    der.avg.water <- mean(c(Der[1,1], Der[1,5], Der[1,9]))
    # Remove background and Normalize the data
    Der <- round(Der * (der.avg.water/anc.avg.water), digits = 3)
    for (m in resources){
      data[which(data$Resource == m), 3 + l] <- mean(Der[which(resource.names == m)])
    }
  }
  data$D1fold <- round(apply(data[, 4:6], 1, mean, na.rm = TRUE) / data$Anc, 3)
  data$D2fold <- round(apply(data[, 7:9], 1, mean, na.rm = TRUE) / data$Anc, 3)
  data$D3fold <- round(apply(data[, 10:12], 1, mean, na.rm = TRUE) / data$Anc, 3)
  for (n in 1:length(resources)){
    if (data$D1fold[n] >= 1){
      data$D1fold[n] <- data$D1fold[n] - 1
      } else {
        data$D1fold[n] <- -(1/data$D1fold[n]) + 1
      }
     if (data$D2fold[n] >= 1){
      data$D2fold[n] <- data$D2fold[n] - 1
      } else {
        data$D2fold[n] <- -(1/data$D2fold[n]) + 1
      }   
     if (data$D3fold[n] >= 1){
      data$D3fold[n] <- data$D3fold[n] - 1
      } else {
        data$D3fold[n] <- -(1/data$D3fold[n]) + 1
      }   
    }
  newdata <- data[order(data$Group),]
  
  data.plot <- newdata[,c(1:2, 13:15)]
  data.plot <- melt(data.plot)
  data.plot <- data.plot[order(data.plot$Group, data.plot$Resource), ]
  
  assign(paste("Strain", i, sep=""), newdata)
  assign(paste("StrainPlot", i, sep=""), data.plot)
}
```

```{r}
### New Data Frame For Plotting
pad <- 4
amine <- 1:6
amino.acid <- (7:24) + pad
carb <- (25:45) + tail(amino.acid, n=1) - 25 + pad + 1
carbox <- (46:72) + tail(carb, n=1) - 46 + pad + 1
ester <- (73:75) + tail(carbox, n=1) - 73 + pad + 1
phos <- (76:81) + tail(ester, n=1) -76 + pad + 1
poly <- (82:93) + tail(phos, n=1) - 82 + pad + 1
x.axis <-c(amine, amino.acid, carb, carbox, ester, phos, poly)

plot.data <- cbind(x.axis, StrainPlot0711)
plot.data[,6] <- NA

plot.data[1:6,6] <- as.character(rep("red", 6))
plot.data[7:24,6] <- as.character(rep("blue", 18))
plot.data[25:45,6] <- as.character(rep("darkgreen", 21))
plot.data[46:72,6] <- as.character(rep("purple", 27))
plot.data[73:75,6] <- as.character(rep("magenta", 3))
plot.data[76:81,6] <- as.character(rep("darkorange", 6))
plot.data[82:93,6] <- as.character(rep("DeepPink4", 12))

#colnames(plot.data[6])[colnames(plot.data[6])=="V6"] <- "color"
```

```{r}
### plotting
plot.new()
png(filename="~/GitHub/SSRP_2015/figures/lollipops.png", width = 1800, 
    height = 800, res = 96*2)

plot(plot.data$x.axis,plot.data$value, type="h", col = plot.data[,6], 
     ylim = c(-2.5, 2.5), yaxt='n', xaxt = 'n', xlab = "", ylab = "", lty = 3,
     panel.first={
       rect(-5, -1, 150, 1, density = NULL, angle = 45, col = "LightGrey", lty = 0)
})

mtext('Fold Change', side = 2, outer = TRUE, cex = 1.5, 
      line = -1.5, adj = 0.55)

axis(side = 2, at = c(-2, -1, 0, 1, 2), labels = c("-3", "-2", "", "2", "-3"), las = 1)
axis(side = 4, at = c(-2, -1, 0, 1, 2), labels = c("", "", "", "", ""), las = 1)
abline(h = 0, lty = 2)

points(plot.data$x.axis,plot.data$value, type = "p", pch = 19, col = plot.data[,6])

text(3, -2, "amines", col = "red", cex = 1, font = 2)
text(19, -2, "amino acids", col = "blue", cex = 1, font = 2)
text(42, -2, "carbohydrates", col = "darkgreen", cex = 1, font = 2)
text(72, -2, "carboxylic acids", col = "purple", cex = 1, font = 2)
text(89, -2, "esters", col = "magenta", cex = 1, font = 2)
text(99, -2, "phosph", col = "darkorange", cex = 1, font =2)
text(112, -2, "polymer", col = "DeepPink4", cex = 1, font =2)

dev.off()
graphics.off()
```
